A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-02 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: more on the section about pseudo-probability

\usepackage{assumptionsofphysics}
\usepackage{tikz}
\usepackage{breakurl}
\usepackage{tcolorbox}
\usepackage{scrextend}	% for environment ``labeling''

@@

\title{A non-additive generalization of probability theory\\for quantum mechanics and beyond}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics, and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math is needed. For the more philosophically inclined, we will point to areas where further conceptual work is needed.

	%Given a generic space of ensembles, we can define the fraction capacity as the maximum fraction of a particular ensemble that can be understood as a mixture of ensembles from a given set. This gives a non-additive (i.e. fuzzy) measure that reduces to a probability (i.e. additive) measure for classical spaces and for quantum measurements. We can also define the state capacity as the exponential of the maximum entropy reachable by a mixture of ensembles from a given set. This is also a non-additive measure and reduces to the Liouville measure in classical mechanics and the Hilbert space dimensions for subspaces of quantum mechanics. Conceptually, it gives us a notion of probability that is both theory and interpretation independent. Mathematically, it gives us a measure theoretic generalization of probability. Physically, it may allow to find new way to understand current theory and tool to investigate new ones. The purpose of this paper is to show the core ideas and present questions that may be developed on the mathematical, physical and philosophical side.

\end{abstract}

@@

\tableofcontents

%\section{TOC}
%
%Overall goal: to create a generalization of probability theory and measure theory for general physical systems.
%
%General question: how do we represent physical states? Different types of representations: talk about examples, ....
%
%-- General problem
%
%For this work: assume we have a locally convex topological vector space V. E is a bounded convex subset of V. Equipped with an entropy. Topology generated by entropic balls?
%
%Separateness and orthogonality
%
%Statistical quantities
%
%Show that they reduce to standard things in classical mechanics and quantum mechanics.
%
%-- Connection to Choquet theory 
%
%Choquet theory - in the case that the closure of the space is compact -> representation through measures. May not be unique
%
%Choquet simplex. Uniqueness of decomposition and separateness/orthogonality.
%
%-- Connection to signed probability
%
%TODO: literature search on use of signed measure in QM. ( e.g. \href{https://arxiv.org/pdf/2302.00118}{this paper})
%
%Characterize signed measures as an affine representation of the ensemble space (i.e. bounded convex subset of a locally convex topological vector space).
%
%Show that the probability is signed if and only if multiple decomposition, non-additive, no Choquet simplex
%
%Signed probability is unique, but does not characterize the convex set (ensemble space). Note: the other measures are defined on the extreme points. Knowing the extreme points means you know the convex space (i.e. take the hull). The non-additive measure does not characterize the extreme points, so it doesn't really characterize the space.
%
%-- Our non-additive measure approach
%
%State capacity
%
%Fraction capacity. Show uniqueness. Also show uniqueness on extreme points.
%
%Fraction capacity as supremum of all possible measures
%
%Additivity of fraction capacity (and state capacity).
%
%Single decomposition and classical spaces and measurement contexts
%
%Show that they reduce to standard things in classical mechanics and quantum mechanics.
%
%-- Connection to fuzzy logic?
%
%Literature search.
%
%Possible connection to
%* Choquet integral
%* Sugeno integral
%
%? Space of fuzzy measure like we have a space of additive measures?

@@

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used, given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.

We reformulate the problem in the following way: given the space of all possible statistical ensembles allowed by a statistical theory, what measure theoretic tools can we use to represent each ensemble and the geometric properties of the overall space? Why exactly do classical probability measures fail to capture completely quantum states? What suitable generalizations are possible, and what mathematical tools would need to be developed?

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurement (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review some results from Choquet theory, which is, to our knowledge, the most general approach to represent elements of a convex set using additive probability measures. The key insight is that, at least in the compact case, such a representation is always possible, but it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex and is a particularly shaped subset of a vector space. The space of quantum ensembles is not a simplex for one key reason: it allows multiple decompositions in terms of pure states. These insights will allow to better understand exactly the limits of classical probability to describe ensembles in physical theories.

We will also review some results from the use of quasi-probability measures. While the Wigner function represents the most known instance, there are different choices. Note that the space of signed probability measures over a set $K$ is a vector space, so one can find an affine map between a suitable space of signed probability measures and the embedding vector space. Each ensemble, then, is uniquely represented with a signed probability measure. However, unlike in the Choquet case, the measure is not supported by the set of extreme points of the ensemble space. That is, there is no unique or privileged choice for this representation. Moreover, since the map is at the level of the full vector space, not just the subset with valid ensembles, there is no general way to characterize which subset of signed probability measures corresponds to ensembles. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of the count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always provide a physically meaningful representation. Moreover, in the compact case, these measures can be restricted to the extreme points while retaining uniqueness. (TODO: prove)

The fraction capacity is the supremum of all possible measures for the same ensemble. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure and is therefore additive. The state capacity, instead, corresponds to the minimum of the entropy of all possible representations. It recovers the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Having shown that a general approach based on non-additive measures is possible, we lay down a series of questions to explore in the future. The biggest one is whether it is possible to generalize notions such as integration, derivation, and expectation to this non-additive case.  It should be noted that non-additive measures are used in the context of decision theory, so there may be a connection between the use of non-additive measures in quantum theory and interdependent complementary choices.[TODO: find citation that sub-additive is when choices are complementary] Both the Choquet and the Sugeno integral, for example, are defined over a monotonic set-functions, so there is a chance that these tools may turn out to be useful. We see this work as starting this line of research, which will take more effort to investigate.

A final remark before we start. Too often, works in the foundations of physics focus on a particular aspect without concern for a broader connection to all relevant mathematical, physical, and philosophical aspects. Therefore, it is important to us that this work is a piece of a much larger puzzle, which connects cleanly and directly to other areas, such as information geometry, functional analysis, quantum logic, and so on. It is the hallmark of a true fundamental structure for physics that such connections can be easily made. Unfortunately, the nature of scientific publication requires us to publish a narrow slice of the work to fit the scope of a particular journal. Ironically, this means that the most important motivations for our approach are out of scope of this article, as we will concentrate only on the connections to measure theory.

@@

Additionally, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.} The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensembles that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when their mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

@@

[This section has to make clear in what extra conditions on ensemble spaces Choquet theory can be applied.]

@@

% Physical meaning of representation with a measure.
First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows us to think of $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choquet simplex by the property of having a single decomposition in terms of extreme points.

% Limitation of applying Choquet theory to our case
We should now look at the requirements of the theorem to see whether we can always apply it to an ensemble space. First of all, an ensemble space $\Ens$ is a convex subset of a convex space $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measures over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore, the theorem does not apply in general.

% Define "Choquet space" as those spaces where we can apply Choquet theory
We can apply Choquet's theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allow a representation through a probability on points that may correspond to idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distributions over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

Note that ensembles over classical spaces will be represented by a unique probability measure, therefore the corresponding closure of the ensemble space is a Choquet simplex. However, quantum mixed states can be decomposed in multiple ways in terms of pure states, and therefore, the closure of a quantum ensemble space is not a simplex.

@@

Chouqet theory tells us, in the realm where it can be applied, that classical probability fails in quantum mechanics precisely because the ensemble space of a quantum system allows multiple decompositions.

@@

A measure $\mu: \Omega_X \to \mathbb{R}$ over a sample space $X$ with $\sigma$-algebra $\Omega_X$ is called a probability measure if it satisfies the following three axioms\footnote{TODO: These should be stated earlier.}
\begin{labeling}{Non-negativity:}
	\item[Non-negativity:] $\forall x \in \Omega_X: \mu(x) \ge 0$
	\item[Normalization:] $\mu(X) = 1$
	\item[$\sigma$-additivity:] $\mu ( \bigcup_i x_i ) = \sum_i \mu ( x_i )$ for any countable sequence of disjoint sets $x_i \in \Omega_X$.
\end{labeling}

@@

\textbf{Not about negative probability.} It is very common to conceive the main difference between classical probability and these quasi-probabilities as the inclusion of negative probability. Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a probability distribution.

The crucial difference is that in classical probability all points of the sample space represent mutually exclusive events. That is, they can be prepared and observed independently from the others. Therefore, any two events that are disjoint are also mutually exclusive. This is not the case for quasi-probability in quantum mechanics. Husimi Q represents a state as decomposed over coherent states, which cannot represent mutually exclusive as they are not orthogonal. Similarly, we can imagine to embed the Bloch ball into a tetrahedron, which means every point in the Bloch ball could be represented as a convex combination of extreme points, meaning that they would be represented by a non-negative quasi-probability distribution over four elements. However, none of the four elements represent a valid state.

@@

For example, affine measures do not necessarily satisfy non-negativity. The Wigner function is the probability density of such a measure and can be negative over small regions of phase space.

Given a space $X$, the space of all affine measures is a vector space, not just a convex subset as in the case of the space of probability measures. This means that, given an ensemble space, we may find a space $X$ such that the space of affine measures over $X$ is isomorphic to the embedding vector space $V$ as topological vector spaces. In this case, any element $v\in V$ can be represented uniquely by an affine measure $\mu_{v}$, regardless of whether they are actual ensembles or not.\footnote{The question of whether such an $X$ always exists is left open. We suspect a link exists between the statistical quantities $F_i$ that fully identify each ensemble and the space $X$.} Because the statistical quantities $F_i$ are affine functions, we will have $F_i(v) = \int_X F_i d\mu_{v}$.

In general, since the ensemble space is bounded along every direction, not all affine measures will correspond to an ensemble space. Whether all probability measures correspond to a valid ensemble, or whether affine measures are even used, corresponds to relationships between $\Ens$ and the space of probability measures $M_p(X)$. If $\Ens \subset M_p(X)$, then all states are represented by a probability measure. However, this is not a ``probability'' in the standard sense of the word. For example, the Husimi Q representation is always non-negative at every point, but two points do not represent mutually exclusive events as they are not orthogonal (i.e. their inner product is non-zero and they do not maximize the entropy increase during mixture). Mathematically, the issue is that the points $X$ of the Husimi Q representation are not the extreme points of the closure ensemble space $\overline{\Ens}$. In general, $\Ens$ and $M_p(X)$ will overlap, but neither will contain the other. For example, in the Wigner representation, the only pure states (i.e. the extreme points of $\Ens$) represented by a probability measure are the Gaussian wave-packets. This means that any other non-negative Wigner function will be in the convex hull of Gaussian wave-packets, and therefore will allow a decomposition in terms of them.

Note that if $\overline{\Ens} = M_p(X)$, then the space of the extreme point is exactly the space of Dirac measures, and therefore $X$ corresponds to the pure states. Alternatively, either one ensemble is represented by an affine measure, or not all probability measures can be understood as mixtures of extreme points. If $\overline{\Ens}$ is compact, then this is exactly when $\overline{\Ens}$ is not a simplex and multiple decompositions are allowed.

@@

This alternative way of looking at measure theoretic representations, then, is consistent with the previous one. While quasiprobability distributions are very useful in practice, we want to stress that there is no single choice for $X$. To solve a specific problem, this may actually be useful as one can choose the most convenient $X$ for the problem. To develop a general theory for physics, however, this is not useful as each representation is not general enough.

@@

The first property readily comes from the range of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures; therefore, they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit. If not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure.

Our setup is very similar to what happens in Choquet theory. The fraction capacity, in fact, is not just defined on the extreme points, but it is defined on the whole space of ensembles. In the classical discrete case, for example, if $A \in \Omega \ni \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $\frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed as a mixture of those distributions. Similarly, the probability measures in Choquet theory are, in general, defined on the whole ensemble space, and then one finds those particular ones that have support on the extreme points.

>>>>

\usepackage{assumptionsofphysics}
\usepackage{tikz}
\usepackage{breakurl}
\usepackage{tcolorbox}

@@

\title{A non-additive generalization of probability theory \\for quantum mechanics and beyond}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}

@@

\begin{abstract}
	We present a physically motivated generalization of probability theory that is suitable for classical mechanics, quantum mechanics and any future physical theory that allows a statistical description. The goal is to put the use of classical and quantum probability in a broader context, and show how the current mathematical structures are likely not suitable to solve the open problems in the foundations of physics. For the more mathematically inclined, we will point to areas where new math or generalization of established math are needed. For the more philosophically incline, we will point to areas where further conceptual work is needed.

	%Given a generic space of ensembles, we can define the fraction capacity as the maximum fraction of a particular ensemble that can be understood as a mixture of ensembles from a given set. This gives a non-additive (i.e. fuzzy) measure that reduces to a probability (i.e. additive) measure for classical spaces and for quantum measurements. We can also define the state capacity as the exponential of the maximum entropy reachable by a mixture of ensembles from a given set. This is also a non-additive measure and reduces to the Liouville measure in classical mechanics and the Hilbert space dimensions for subspaces of quantum mechanics. Conceptually, it gives us a notion of probability that is both theory and interpretation independent. Mathematically, it gives us a measure theoretic generalization of probability. Physically, it may allow to find new way to understand current theory and tool to investigate new ones. The purpose of this paper is to show the core ideas and present questions that may be developed on the mathematical, physical and philosophical side.
\end{abstract}

@@

\section{TOC}

Overall goal: to create a generalization of probability theory and measure theory for general physical systems.

General question: how do we represent physical states? Different types of representations: talk about examples, ....

-- General problem

For this work: assume we have a locally convex topological vector space V. E is a bounded convex subset of V. Equipped with an entropy. Topology generated by entropic balls?

Separateness and orthogonality

Statistical quantities

Show that they reduce to standard things in classical mechanics and quantum mechanics.

-- Connection to Choquet theory

Choquet theory - in the case that the closure of the space is compact -> representation through measures. May not be unique

Choquet simplex. Uniqueness of decomposition and separateness/orthogonality.

-- Connection to signed probability

TODO: literature search on use of signed measure in QM. ( e.g. \href{https://arxiv.org/pdf/2302.00118}{this paper})

Characterize signed measures as an affine representation of the ensemble space (i.e. bounded convex subset of a locally convex topological vector space).

Show that the probability is signed if and only if multiple decomposition, non-additive, no Choquet simplex

Signed probability is unique, but does not characterize the convex set (ensemble space). Note: the other measures are defined on the extreme points. Knowing the extreme points means you know the convex space (i.e. take the hull). The non-additive measure does not characterize the extreme points, so it doesn't really characterize the space.

-- Our non-additive measure approach

State capacity

Fraction capacity. Show uniqueness. Also show uniqueness on extreme points.

Fraction capacity as supremum of all possible measures

Additivity of fraction capacity (and state capacity).

Single decomposition and classical spaces and measurement contexts

Show that they reduce to standard things in classical mechanics and quantum mechanics.

-- Connection to fuzzy logic?

Literature search.

Possible connection to
* Choquet integral
* Sugeno integral

? Space of fuzzy measure like we have a space of additive measures?

@@

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics.

We reformulate the problem in the following way: given the space of all possible statistical ensembles allowed by a statistical theory, what measure theoretic tools can we use to represent each ensemble and the geometric properties of the overall space? Why exactly classical probability measures fail to capture completely quantum states? What suitable generalizations are possible, and what mathematical tools would need to be developed?

In this paper, we will characterize a generic space of ensembles as a convex bounded set of a Hausdorff locally convex topological vector space, upon which an entropy function is defined. These mathematical properties are physically grounded in the ability to experimentally define ensemble (leading to a topological space), to define statistical mixtures (leading to a convex space), to quantify the variability of the elements of the ensembles (leading to an entropy function) and to having enough real valued quantities to identify each ensemble (leading to the continuous embedding into a Hausdorff locally convex topological vector space). The question is, then, what measure theoretic tool can quantify the number of states (as in the classical Liouville measure) or the number of possible outcomes of a measurements (as in the dimension of orthogonal subspaces of an operator)? What measure theoretic tool can represent each ensemble? If we are to define a physically meaningful notion of probability density, both of these questions must be answered.

We will first review some results from Choquet theory, which is, to our knowledge, the most general approach to represent elements of a convex set using additive probability measures. The key insight is that, at least in the compact case, such representation is always possible, but it is not unique. Uniqueness requires the space to be a Choquet simplex, which is a generalization of the standard finite-dimensional simplex and is a particularly shaped subset of a vector space. The space of quantum ensembles is not a simplex for one key reason: it allows multiple decompositions in terms of pure states. These insights will allow to better understand exactly the limits of classical probability to describe ensembles in physical theories.

We will also review some results from the use of quasi-probability measures. While the Wigner function represents the most known instance, there are different choices. Note that the space of signed probability measures over a set $K$ is a vector space, so one can find an affine map between a suitable space of signed probability measures and the embedding vector space. Each ensemble, then, is uniquely represented with a signed probability measure. However, unlike in the Choquet case, the measure is not supported by the set of extreme points of the ensemble space. That is, there is no unique or privileged choice for this representation. Moreover, since the map is at the level of the full vector space, not just the subset with valid ensembles, there is no general way to characterize which subset of signed probability measures correspond to ensembles. Therefore, while signed measures are a powerful tool for calculation in specific problems, they cannot give us a general principled representation of the physical theory.

The above two approaches exhaust what can be done with additive measures, so we turn to the non-additive case. From the entropy, we can define the state capacity, which generalizes the notion of count of states. For each ensemble, we can define the fraction capacity, which generalizes the notion of probability. Both are non-negative sub-additive monotonic set functions, with the second being unit bounded. These measures are always unique and, since they are defined on the ensemble space itself, always have provide a physically meaningful representation. Moreover, in the compact case, these measure can be restricted on the extreme points while retaining uniqueness. (TODO: prove)

The fraction capacity is the supremum of all possible measures for the same ensemble. The classical case exactly corresponds to the simplex case, in which fraction capacity corresponds to a unique measure, and is therefore additive. The state capacity, instead, corresponds to the minimum of the entropy of all possible representations. It recovers the Liouville measure in the classical case and the dimensionality of the Hilbert space in the quantum case.

Having shown that a general approach based on non-additive measures is possible, we lay down a series of questions to explore in the future. The biggest one is whether it is possible to generalize notions such as integration, derivation and expectation to this non-additive case.  It should be noted that non-additive measures are used in the context of decision theory, so there may be a connection between the use of non-additive measures in quantum theory and interdependent complimentary choices.[TODO: find citation that sub-additive is when choices are complementary] Both the Choquet and the Sugeno integral, for example, are defined over a monotonic set-functions, so there is a chance that these tools may turn out to be useful. We see this work as starting this line of research, which will take more effort to investigate.

A final remark before we start. Too often works in the foundations of physics focus on a particular aspect without concern to a broader connection to all relevant mathematical, physical and philosophical aspects. Therefore, it is important to us that this work is a piece of a much larger puzzle, which connects cleanly and directly to  other areas, such as information geometry, functional analysis, quantum logic and so on. It is the hallmark of a true fundamental structure for physics that such connections can be easily made. Unfortunately, the nature of scientific publication requires us to publish a narrow slice of the work to fit the scope of a particular journal. Ironically, this means that the most important motivations for our approach are out of scope of this article, as we will concentrate only to the connections to measure theory.

@@

Additional, each ensemble space $\Ens$ is equipped with a continuous entropy function $S : \Ens \to \mathbb{R}$. The entropy of a statistical mixture satisfies the following bounds
\begin{equation}
	\begin{aligned}
		p S(\ens[a]) + (1-p) &S(\ens[b]) \leq S(p \ens[a] + (1-p) \ens[b]) \\
		&\leq I(p,\bar{p}) + p S(\ens[a]) + (1-p) S(\ens[b])
	\end{aligned}
\end{equation}
where $I(p,\bar{p}) = - p \log p - \bar{p} \log(\bar{p})$.\footnote{In the larger work, we can show that the Shannon entropy is the only upper bound possible.} The bounds come from assuming that the entropy quantifies the variability of each ensemble. Intuitively, the variability cannot decrease during mixing, and will have an upper bound when mixing two ensembles that are non-overlapping, or orthogonal. In fact, we define two ensembles $\ens[a] \ortho \ens[b]$ to be orthogonal when they mixture gives a maximal entropy increase. This does recover the notion of orthogonality in the corresponding classical and quantum vector spaces of statistical ensembles. Lastly, mixtures preserve orthogonality in the sense that mixing non-overlapping ensembles still gives non-overlapping ensembles. More precisely, $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$.

@@

[This section has to make clear in what extra conditions on ensembles spaces Choquet theory can be applied.]

@@

% Physical meaning of representation with a measure.
First, let us understand the physical significance. An ensemble $\ens$ can be seen as a probability measure $\mu$ over the ``pure states'' (i.e. the extreme points of $E$) if the value for every statistical quantity $f$ for $\ens$ can be recovered as an expectation over the pure states using the probability measure. That is, all statistical properties of $\ens$ can be recovered by the probability measure. This is what allows to think go $\ens$ as a probability distribution. Note, however, that the Choquet theorem does not guarantee uniqueness. The probability measure is unique if and only if $E$ is a Choquet simplex, which is a generalization of the finite dimensional simplex. The characterization of the Choquet simplex is fairly technical, so we will simply define the Choque simplex by the property of having a single decomposition in terms of extreme points.

% Limitation of applying Choquet theory to our case
We should now look at the requirements of the theorem, to see whether we can always apply it to an ensemble space. First of all, and ensemble space $\Ens$ is a convex subset of a convex sapce $V$. Moreover, since every Hausdorff first-countable topological vector space is metrizable, $\Ens$ is metrizable. However, $\Ens$ is not necessarily compact. In fact, $\Ens$ is not necessarily a closed subset of $V$. Consider the space of probability measures over $\mathbb{R}$. Since $\mathbb{R}$ is not compact, the space of probability measure over $\mathbb{R}$ cannot be compact.\footnote{TODO: Check with mathematicians} Moreover, the set of extreme points is given by the Dirac measures (i.e. probability one over a single point), which would have minus infinite entropy, and therefore cannot be part of the ensemble space. Therefore the theorem does not apply in general.

% Define "Choquet space" as those spaces where we can apply Choquet theory
We can apply Choquet theorem in the case where the closure of $\Ens$ in $V$ is a compact set $E$, which will also be convex\footnote{The closure of a convex set is convex. TODO: finalize the proof} and metrizable. In this case, we are guaranteed that all ensembles allows a representation through a probability on points that may corresponds idealized ensembles (i.e. the Dirac measure) that are not, even conceptually, realizable. Many interesting cases fall into this restriction: the ensemble spaces of classical probability distribution over finitely many discrete cases fall in this restriction; the subset of ensembles over finite ranges of position and momentum; the ensemble spaces of all finite dimensional quantum systems.

Note that ensembles over classical spaces will be represented by a unique probability measure, therefore the corresponding closure of the ensemble space is a Choquet simplex. However, quantum mixed states can be decomposed in multiple ways in terms of pure states, and therefore the closure of a quantum ensemble space is not a simplex.

@@

Chouqet theory tells, in the realm where it can be applied, that classical probability fails in quantum mechanics precisely because the ensemble space of a quantum system allows multiple decomposition.

@@

@@

\textbf{Not about negative probability.} It is very common to conceive the main difference between classical probability and these quasi-probabilities as the inclusion of negative probability. Sometimes these negative probabilities are justified as arising from constraints on probability distributions defined on different contexts. This is a mischaracterization. For example, the Husimi Q representation provides us with a non-negative quasi-probability function. That is, the Husimi Q is a non-negative density distribution that integrates to one, yet it is not a classical probability distribution. Similarly, since we can embed the Bloch ball inside a tetrahedron, we can represent all states of a two-state quantum system by a convex combination of the vertices, leading to a non-negative quasi-probability distribution over four elements.

Conversely, we can use signed measure to represent classical distributions. Suppose we have a two state classical discrete space, which means the space of all possible ensembles is spanned by $[p, (1-p)]$. Alternatively, we can write it as $p[1,0] + (1-p)[0,1]$ which highlights that we are writing each possible probability distribution in terms of the two possible cases. But we could also express each probability distribution in terms of other two distributions, as in $\left(2p - \frac{1}{2}\right)\left[\frac{3}{4}, \frac{1}{4}\right] + \left(\frac{3}{2} - 2p\right)\left[\frac{1}{4}, \frac{3}{4}\right]$. The coefficients now span between $\left[-\frac{1}{2}, \frac{3}{2}\right]$, can be negative, but still sum to one. That is, in terms of this new affine basis, we need to use a signed probability.

Since we can represent classical states with signed probability measures, and quantum states with non-negative probability measure, the difference between classical probability and quantum quasi-probability does not lie in the use of negative probability. Instead, the crucial difference is that in classical probability all points of the sample space, and therefore disjoint regions, represent mutually exclusive events. That is, they can be prepared and observed independently from the others; we can assign some probability to some region without having to assign it to another disjoint region.\footnote{The only constraint is that the probability over the whole space is one.} This is not the case for quasi-probability in quantum mechanics. Husimi Q represents a state as decomposed over coherent states, which cannot represent mutually exclusive as they are not orthogonal. The inner product between these states will provide a constrain. In the non-negative probability representation of the Bloch sphere, none of the four corners represent valid states, as they lie outside the Bloch ball. So we cannot assign probability to a vertex without assigning some probability to another vertex.

While negative probability is a red herring, we want to understand two things. When can we use classical probability? Are quasi-probability representation specific to quantum mechanics, or are they more general? Fortunately, the ensemble space framework allows us to answer these questions directly.

@@

\textbf{Ensemble space embedding and signed measures.} We have already seen that we can represent ensembles using classical probability if and only the ensemble space is a simplex. Since any probability measure is possible, we can assign probability to two disjoint areas independently. This answers the first question.

For the second question, let $Z$ be the space over which we are defining our quasi-probability distributions.
Note that, in general, $Z$ is not the space of states. Wigner functions, for example, are defined over phase space, which does not represent quantum states. Let $M^{\pm}(Z)$ the space of all bounded signed measures over $Z$. Since the linear combination of two bounded signed measure is a bounded signed measure, $M^{\pm}(Z)$ is a vector space. Now let $M_1^{\pm}(Z)$ be the space of signed probability measures. While it is not a subspace of $M(Z)$, it does not include the zero measure, it is an affine subspace therefore, up to the choice of an arbitrary origin, it is a vector space.

Recall that an ensemble space must be a convex subset of a vector space. If we are able to represent ensembles as quasi-probability distribution over $Z$, then, $M_1^{\pm}(Z)$ will be isomorphic to the embedding vector space $V$ as topological vector spaces.\footnote{The question of whether such an $Z$ always exists is still open for us. A possible conjecture is that $Z$ could be constructed from the statistical quantities $F_i$ that fully identify each ensemble in $\Ens$.} In this case, any element $v\in V$ can be represented uniquely by a signed probability measure $\mu_{v}$, regardless of whether it is an actual ensemble or not. Since statistical quantities $F_i$ are linear functional, we will have $F_i(v) = \int_Z F_i(z) d\mu_{v}$. In other words, there is nothing special about quantum mechanics that allows quasi-probability representations. It is the fact that the ensemble space embeds into a vector space, a general fact for all theories, that allows such representations.

It should be clear that, since an ensemble space is a strict subset of the embedding vector space, not all affine measure will correspond to an ensemble space. Whether all probability measures correspond to a valid ensemble, or whether signed probability measures are even used depends on how the mapping is done in each particular case. For example, there is no simple characterization of which Wigner functions correspond to valid quantum states and which do not.

@@

\textbf{Non-uniquess of quasi-probability.}

This alternative way of looking at measure theoretic representations, then, is consistent with the previous one and fits in our general framework. Since quantum ensemble spaces are not simplexes, classical probability does not apply. The different bounds one gets between classical and quantum mechanics depend on the different shape of the respective ensemble spaces.

Quasi-probability distributions are very useful in practice, particularly because one can construct them to fit particular problems. Once one has identified a set of observables that are the most natural ones to define a problem, one can find a representation that is specifically suited for the case. While the quasi-probability cannot be interpreted as probability, it still provides a useful tool to describe and understand the problem.

However, this practical convenience is also a drawback for finding a more general approach to probability in spaces beyond classical mechanics. Crucially, the space $Z$ over which the quasi-probability measure are defined is not unique, and it is not defined by the structure of the ensemble space and, as a consequence, cannot tells us which quasi-probability measure represent valid ensembles and which do not.

@@

The first property readily comes from the range of the fraction. For the second, note that $\hull$ and $\sup$ are both monotone. For sub-additivity, note that the fraction of two components can at most sum during mixing. For the last, increasing and decreasing sequences strictly add or remove possible mixtures, therefore they will lead to increasing or decreasing sequences of real numbers bounded between zero and one. The limit of these sequences must agree with the fraction capacity of the limit. If not for the lack of additivity over disjoint sets, the fraction capacity would be a probability measure.

Our setup is very similar to what happens in Choquet theory. The fraction capacity, in fact, is not just defined on the extreme points, but it is defined on the whole space of ensembles. In the classical discrete case, for example, if $A \subset \Omega \subset \Ens$ is an event, then $\frcap_{\ens}(A)$ will return the probability of event $A$; if $P \subset \Ens$ represents all the Poisson distributions with integer coefficient, $frcap_{\ens}(P)$ will return how much of $\ens$ can be expressed is a mixture of those distributions. Similarly, the probability measures in Choquet theory are, in general, defined on the whole ensemble space, and then one finds those particular ones that have support on the extreme points.
